Ecosystem size and complexity as extrinsic drivers of food chain length in branching ecosystems

### 2023-03-03 rmd/manuscript.Rmd
edit: manuscript

Among the three hypotheses, the ecosystem size hypothesis has been consistently supported in empirical explorations, likely because it encompasses multiple mechanisms [@takimotoEnvironmentalDeterminantsFoodchain2013]. For example, large ecosystems have increased species richness (species-area relationship [@lomolinoEcologyMostGeneral2000]), are buffered against disturbances [@saboRoleDischargeVariation2010], and have greater numbers of immigrant sources in the ecosystem with stable metapopulations [@takimotoEnvironmentalDeterminantsFoodchain2013]. Meanwhile, the theoretical basis for the other two hypotheses rests on implicit assumptions of linear food chains, which are often violated in natural systems. As such, the effects of disturbance and resource availability on FCL have been found to be highly variable among ecosystem types and geographic regions [@takimotoEffectsProductivityDisturbance2012; @diehlIntraguildPreySuffer2001; @takimotoEnvironmentalDeterminantsFoodchain2013; @postLongShortFoodchain2002].

@@

@@

Another interesting result of our simulation is the interaction with omnivory. In particular, when the chain motif was combined with the no-disturbance scenario, ecosystem size and complexity had contrasting influences on FCL. It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity; therefore, the total carrying capacity within a network dictates FCL. The effect of ecosystem size is understandable because it simply increases the total carrying capacity. In contrast, ecosystem complexity, here assessed as branching probability, decreases the total carrying capacity as more branched networks have disproportionate numbers of headwater habitats with small local carrying capacities. However, the dependence of FCL on local carrying capacity diminished in weak and strong omnivory scenarios, explaining the lack of positive (ecosystem size) or negative FCL-geomtery relationship (ecosystem complexity) in these motifs with no disturbance.

@@

>>>>

Among the three hypotheses, the ecosystem size hypothesis has been consistently supported in empirical explorations, likely because it encompasses multiple mechanisms [@takimotoEnvironmentalDeterminantsFoodchain2013]. For example, large ecosystems have increased species richness (species-area relationship [@lomolinoEcologyMostGeneral2000]), are buffered against disturbances [@saboRoleDischargeVariation2010], and have greater numbers of immigrant sources in the ecosystem with stable metapopulations [@takimotoEnvironmentalDeterminantsFoodchain2013]. Meanwhile, the productivity and disturbance effects are highly sensitive to the internal structure of food web motifs. As such, the effects of disturbance and resource availability on FCL have been found to be highly variable among ecosystem types and geographic regions [@takimotoEffectsProductivityDisturbance2012; @diehlIntraguildPreySuffer2001; @takimotoEnvironmentalDeterminantsFoodchain2013; @postLongShortFoodchain2002].

@@

State transition shed lights on how FCL changes occurred. In the chain motif, the "additive" mechanism (i.e., the addition of higher trophic levels) is always responsible for the changes in FCL. With disturbance, state transition occurred from "B" to "B + C" as ecosystem size or complexity increases. With no disturbance, state transition was different between size and complexity. Ecosystem size facilitated the transition from "B + C" to "B + C + P", while ecosystem complexity drove the opposite transition ("B + C + P" $\rightarrow$ "B + C").

In the weak and strong omnivory motifs, ecosystem size and complexity increased FCL through different mechanisms. Ecosystem size increased FCL primarily through the additive mechanism, facilitating the state transition from "B" to "B + P", particularly with frequent disturbance. However, the "insertion" mechanism (the transition from "B + P" to "B + C + P") also took place in the strong omnivory motif with rare disturbance ($p_m = 0.05$). In contrast, the insertion mechanism was responsible for the effect of ecosystem complexity on FCL. In the weak omnivory scenario, the transition from "B + P" to "B + C + P" was predominant. In the strong omnivory scenario, transition patterns depended on disturbance frequency. When disturbance was rare ($p_m = 0.05$), the insertion mechanism ("B + P" $\rightarrow$ "B + C + P") drove the FCL change. However, as disturbance frequency increased, the primary process switched to the additive mechanism ("B" $\rightarrow$ "B + P").

@@

Another interesting result of our simulation is the interaction with omnivory. In particular, when the chain motif was combined with no-disturbance, ecosystem size and complexity had contrasting influences on FCL. It is evident from the one-patch simulation that the chain motif is highly sensitive to local carrying capacity; therefore, the total carrying capacity within a network dictates FCL. The effect of ecosystem size is understandable because it simply increases the total carrying capacity. In contrast, ecosystem complexity, here assessed as branching probability, decreases the total carrying capacity as more branched networks have disproportionate numbers of headwater habitats with small local carrying capacities. However, the dependence of FCL on local carrying capacity diminished in weak and strong omnivory scenarios, explaining the lack of positive (ecosystem size) or negative FCL-geomtery relationship (ecosystem complexity) in these motifs with no disturbance.

@@

The effect of productivity was context dependent, supporting previous theoretical predictions. While productivity clearly increased FCL in the chain motif, this effect was vague (strong omnivory) or even reversed (weak omnivory). High productivity effectively increases the abundance of the basal species, stimulating the prey switching of the intraguild predator from the intraguild prey to the basal species. As such, increased productivity does not necessarily increase FCL. Our results suggest that this prediction from non-spatial models can be applied to spatially complex ecosystems.

Ecosystem size and complexity led to the distinct patterns of state transitions. The state transitions with increased ecosystem size are likely associated with increased immigration. It is clear that the total number of dispersing propagules increases with increasing ecosystem size, providing opportunities to maintain longer food chains. The exception was the strong omnivory motif with rare disturbance, in which the insertion mechanism operated. This is likely because rare disturbance inhibited the exclusion of the intraguild prey by suppressing the abundance of the intraguild predator. Similarly, increased immigration may be generally responsible for the state transitions with increasing ecosystem complexity. Habitat patches are closer in more branched networks as it compresses the network configuration. However, unlike ecosystem size, ecosystem complexity does not increase the total number of dispersing propagules - rather, it enhances connectivity among habitat patches. This difference may explain the distinct patterns of state transitions along the gradients of ecosystem size and complexity.
